---
title: "Capped colour scales in detail"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Capped colour scales in detail}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.6)
```

```{r}
#| message: false
library(airquality.methods)
library(ggplot2)
```

`vignette("scales")` introduces the capped scale in one figure. This vignette goes
through its options one by one: one-sided caps, quantile limits, palettes, binned and
discrete variants, the whisker, and the way back to ordinary ggplot2 behaviour.

## Test data

A raster field with a few strong outliers at the top and at the bottom, so that every
example has something to cap.

```{r}
set.seed(42)
d <- expand.grid(x = 1:60, y = 1:40)
d$z <- 50 + 35 * sin(d$x / 7) * cos(d$y / 5) + rnorm(nrow(d), 0, 6)
d$z[d$x < 4  & d$y < 4]  <- -40      # outlier below
d$z[d$x > 57 & d$y > 37] <- 210      # outlier above

range(d$z)

ras <- function(...) {
  ggplot(d, aes(x, y, fill = z)) + geom_raster() +
    coord_fixed() + theme_bw() + labs(x = NULL, y = NULL)
}
```

## The default case

Capped at 0 and 100: values below and above get the edge colour. The edge class is
thus "x ≤ 0" or "x ≥ 100" — the limit itself belongs to it, because `oob_squish`
leaves it in place and pulls everything beyond it there. The legend is labelled
accordingly, and a whisker beside the bar shows how far the data actually reach.

```{r}
ras() + scale_fill_capped(limits = c(0, 100))
```

## Capping one side only

`NA` leaves that side open — then there is neither "≤"/"≥" nor a whisker there.

```{r}
ras() + scale_fill_capped(limits = c(0, NA))
```

## Limits as quantiles

`probs` is computed across all data mapped to this scale, including all facets.

```{r}
ras() + scale_fill_capped(probs = c(0.02, 0.98))
```

A single number stands for `c(p, 1 - p)`; `NA` caps only one end:

```{r}
ras() + scale_fill_capped(probs = 0.05)
ras() + scale_fill_capped(probs = c(NA, 0.98))
```

`limits` and `probs` are mutually exclusive: fixed limits, or limits from the data —
not both.

## Palettes

A name from `grDevices::hcl.pals()`:

```{r}
ras() + scale_fill_capped(limits = c(0, 100), palette = "Zissou 1")
```

A colour vector, here with `values` to shift the midpoint of a diverging palette:

```{r}
ras() + scale_fill_capped(
  limits  = c(0, 100),
  palette = c("#2166AC", "#F7F7F7", "#B2182B"),
  values  = c(0, 0.35, 1)
)
```

A palette function `f(n)`, and a reversed palette:

```{r}
ras() + scale_fill_capped(limits = c(0, 100), palette = scales::pal_viridis(option = "A"))
ras() + scale_fill_capped(limits = c(0, 100), palette = "YlOrRd", direction = -1)
```

## Binned

```{r}
ras() + scale_fill_capped_b(limits = c(0, 100), n.breaks = 5)
```

With custom class boundaries, a palette with exactly as many colours as classes is used
as it is, not interpolated:

```{r}
ras() + scale_fill_capped_b(
  limits  = c(0, 100),
  breaks  = c(20, 40, 60, 80),
  palette = c("#2C7BB6", "#ABD9E9", "#FFFFBF", "#FDAE61", "#D7191C")
)
```

## Discrete

Capping and whiskers mean nothing for categories; the discrete variant is the shared
frame around the same palette logic, so a set of figures can use one palette source.

```{r}
#| fig.height: 3
dd <- data.frame(g = LETTERS[1:5], v = c(3, 5, 2, 8, 4))
ggplot(dd, aes(g, v, fill = g)) + geom_col() +
  scale_fill_capped_d(palette = "Zissou 1") + theme_bw()
```

## Colour instead of fill, and several aesthetics

```{r}
pts <- d[d$x %% 3 == 0 & d$y %% 3 == 0, ]

ggplot(pts, aes(x, y, colour = z)) +
  geom_point(size = 2) +
  scale_colour_capped(limits = c(0, 100)) + theme_bw()

# both aesthetics on one scale
ggplot(pts, aes(x, y, colour = z, fill = z)) +
  geom_point(size = 3, shape = 21) +
  scale_capped(aesthetics = c("colour", "fill"), limits = c(0, 100)) + theme_bw()
```

## Horizontal legend

```{r}
#| fig.height: 4
ras() + scale_fill_capped(limits = c(0, 100)) +
  theme(legend.position = "bottom")
```

## The whisker

The whisker is built like a boxplot's: a stem from the bar edge outward, and a cross
stroke marking the extreme value.

```{r}
ras() + scale_fill_capped(
  limits            = c(0, 100),
  whisker_gap       = unit(4, "mm"),                       # fixed gap
  whisker_width     = 0.6,                                 # fraction of bar width
  whisker_colour    = "grey30",
  whisker_linewidth = 0.8,
  whisker_labels    = scales::label_number(accuracy = 1, suffix = " ug/m3")
)
```

No stem — just the cross stroke, free-standing:

```{r}
ras() + scale_fill_capped(limits = c(0, 100), whisker_stem = FALSE)
```

Only "≤"/"≥" without whiskers, or only whiskers without "≤"/"≥":

```{r}
ras() + scale_fill_capped(limits = c(0, 100), whiskers = FALSE)
ras() + scale_fill_capped(limits = c(0, 100), cap = FALSE)
```

The shorthand "<"/">" instead of the accurate "≤"/"≥". The limit belongs to the edge
class (`oob_squish` leaves it in place and pulls everything beyond it there), so
`include_cap_val = FALSE` drops that information:

```{r}
ras() + scale_fill_capped(limits = c(0, 100), include_cap_val = FALSE)
```

## Arguments passed through to ggplot2

`name`, `breaks`, `labels` and the guide's own settings work as in any ggplot2 scale:

```{r}
#| fig.height: 4.5
ras() + scale_fill_capped(
  limits = c(0, 100),
  name   = "NO2\n(ug/m3)",
  breaks = seq(0, 100, 20),
  labels = scales::label_number(big.mark = "'"),
  guide  = guide_colourbar_capped(
    theme = theme(legend.key.height = unit(3, "cm")),
    reverse = FALSE
  )
)
```

On a transformed scale the cap and whisker labels are back-transformed:

```{r}
dl <- transform(d, z = 10^((z + 50) / 60))
ggplot(dl, aes(x, y, fill = z)) + geom_raster() + coord_fixed() + theme_bw() +
  scale_fill_capped(limits = c(1, 1000), transform = "log10")
```

## Facets

Quantile limits are computed jointly across all facets, so every panel uses the same
colours for the same values:

```{r}
#| fig.width: 7
df <- rbind(transform(d, f = "North"), transform(d, f = "South", z = d$z * 1.4))
ggplot(df, aes(x, y, fill = z)) + geom_raster() + facet_wrap(~f) +
  scale_fill_capped(probs = c(0.02, 0.98), palette = "Batlow") +
  coord_fixed() + theme_bw()
```

## Back to ggplot2's default behaviour

`guide = "colourbar"` switches off the cap labels and whiskers; the colour capping
(`oob = squish`) stays:

```{r}
ras() + scale_fill_capped(limits = c(0, 100), guide = "colourbar")
```

`oob = scales::oob_censor` makes it an entirely ordinary scale: values outside the
limits become `na.value`. Then there is no edge class left, and "≤"/"≥" and the
whiskers disappear on their own — they would otherwise claim something false.

```{r}
ras() + scale_fill_capped(limits = c(0, 100), oob = scales::oob_censor)
```
