---
title: "Reading geodata from data.geo.admin.ch"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Reading geodata from data.geo.admin.ch}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", eval = FALSE)
```

Every chunk below talks to the STAC API of data.geo.admin.ch, so nothing is
evaluated when the vignette is built. Run it interactively; downloads are cached
under `geo_admin_cache_dir()`, so a second run is offline.

```{r}
#| eval: true
#| message: false
library(airquality.methods)
library(dplyr)
library(stars)
```

## The shape of the problem

Air quality exposure combines datasets that do not agree on anything:

| Dataset | Resolution | Years | Quantity |
|---|---|---|---|
| STATPOP inhabitants | 100 m | 2010– | count |
| NO2, PM10, O3 | varies, higher from 2020 | 1990– | concentration |
| PM2.5 | as above | 2015– | concentration |
| Nitrogen deposition | 500 m | 1990, 2000, 2005, 2010, 2015, 2020 | rate |

Different grids, different years, and quantities that must not be resampled the
same way. The workflow below keeps each of those differences explicit.

## 1. Read, in the original resolution

`read_geo_admin()` queries the assets of each collection, resolves the file
hierarchy down to one file per year, and reads it. Nothing is resampled yet.

```{r}
collections <- c(
  "ch.bfs.statistik-bevoelkerung_haushalte",
  "ch.bafu.luftreinhaltung-feinstaub_pm2_5",
  "ch.bafu.luftreinhaltung-stickstoffdeposition"
)

data <- read_geo_admin(collections, years = 2015:2025)
data |> select(label, year, format, grid, noloc_subtracted)
```

The result is one row per collection and year, with the raster in a `stars` list
column and its provenance beside it: `item`, `href`, `format`, `grid`, `res_x`,
`res_y`. Filtering, joining and selecting are plain dplyr from here on.

Years that a collection does not offer are reported rather than silently
dropped — nitrogen deposition has six model years, not eleven.

### Looking at one collection first

When a collection is new or its file layout is unclear, go one level down:

```{r}
assets <- get_geo_admin_assets("ch.bafu.luftreinhaltung-feinstaub_pm2_5")
count(assets, year, format)

# which file would be chosen per year, and why
resolve_assets(assets, formats = "tif", years = 2020:2024)
```

## 2. Stack into cubes, losslessly

```{r}
stack_years(data) |> select(label, grid, years)
```

A `stars` cube has exactly one resolution per dimension, so this yields one cube
per collection *and grid*. If a collection changed resolution between years you
get several cubes — which is the honest answer, not a defect.

## 3. Normalise onto a common grid

Two ways, depending on what the analysis is anchored to.

### 3a. Onto the grid of a reference collection

```{r}
aligned <- align_to_reference(data)
```

For each year of the reference (STATPOP by default) its own raster becomes the
template. The reference is never resampled; everything else is brought onto it.

### 3b. Onto a grid you choose

```{r}
grid_250 <- make_reference_grid(bbox_zh_lv95, cellsize = 250)
normalized <- align_to_grid(data, grid_250)

# or an existing raster as the template:
# align_to_grid(data, "path/to/reference.tif")
```

### How each quantity is resampled

Declared once per collection in `geo_admin_specs()`:

* **`"sum"`** for counts — inhabitants. The total survives coarsening *and*
  refining.
* **`"average"`** for concentrations and rates — area-weighted.

Summing a concentration or averaging a population would both be wrong, so the
choice is part of the collection description rather than a call-site argument.

`sources` records what happened, per year and attribute:

```{r}
normalized$sources[[1]]
```

`total_source` and `total_aligned` show the population total before and after
resampling. They differ only where the target grid is smaller than the data that
was read — which is exactly when you want to know.

### Filling the years between model years

Nitrogen deposition exists for 1990, 2000, 2005, 2010, 2015, 2020. To place it
on an annual axis, say so explicitly:

```{r}
normalized_filled <- align_to_grid(
  data, grid_250,
  years = 2015:2025,
  year_match = c("ch.bafu.luftreinhaltung-stickstoffdeposition" = "linear"),
  max_year_diff = 5
)
```

`year_match` is `"exact"` by default: no data of a different year sneaks in
unannounced. `"nearest"` takes the closest year, `"linear"` interpolates between
the years either side. Neither ever extrapolates — beyond the ends of the series
the nearest year is carried, and `sources` says so.

```{r}
# nitrogen deposition alone, as an annual series on its own 500 m grid
n_dep <- read_geo_admin("ch.bafu.luftreinhaltung-stickstoffdeposition")
n_dep_annual <- align_to_grid(
  n_dep, grid = n_dep$stars[[1]], years = 2000:2020, year_match = "linear"
)
n_dep_annual$sources[[3]] # 2002: 60 % of 2000, 40 % of 2005
```

## 4. One cube

```{r}
cube <- stack_years(normalized)$cube[[1]]
cube
```

Dimensions `x`, `y`, `year`; attributes `BBTOT`, `pm2_5`, `n_deposition`.

## 5. Evaluate

### On the cube directly

Cell-wise across attributes, all years at once:

```{r}
exposure <- cube |>
  mutate(
    pm2_5_exposure = pm2_5 * BBTOT,               # inhabitants x µg/m³
    above_10       = pm2_5 > 10,
    pop_above_10   = ifelse(pm2_5 > 10, BBTOT, 0)
  ) |>
  select(BBTOT, pm2_5_exposure, above_10, pop_above_10)

slice(cube, year, 1)          # one year by index
cube |> filter(year >= 2020)  # by value; needs the cubelyr package

years <- st_get_dimension_values(cube, "year")
change <- slice(cube, year, length(years)) - slice(cube, year, 1)
plot(change["pm2_5"])
```

### Reducing over dimensions

```{r}
st_apply(cube["pm2_5"], c("x", "y"), mean, na.rm = TRUE) # per cell, over years
st_apply(cube["BBTOT"], "year", sum, na.rm = TRUE)       # per year, over the extent

# per municipality, with sf polygons
# aggregate(cube["BBTOT"], by = municipalities, FUN = sum, na.rm = TRUE)
```

### The full dplyr vocabulary, via the long table

```{r}
cube_tbl <- as_tibble(cube) # x, y (cell centres), year, and the attributes

cube_tbl |>
  filter(!is.na(BBTOT), !is.na(pm2_5)) |>
  summarise(
    pop               = sum(BBTOT),
    pm2_5_popweighted = weighted.mean(pm2_5, BBTOT),
    share_above_10    = sum(BBTOT[pm2_5 > 10]) / pop,
    .by = year
  )
```

`tibble_to_cube()` is the way back. It matches cells by the template's
coordinates, so rows removed by a `filter()` simply become `NA`, and the crs,
the grid and years-as-instants are preserved:

```{r}
pm2_5_change <- cube_tbl |>
  arrange(year) |>
  mutate(pm2_5_change = pm2_5 - lag(pm2_5), .by = c(x, y)) |>
  select(x, y, year, pm2_5_change) |>
  tibble_to_cube(cube)

# a summary per cell drops the year dimension and yields a 2D cube
cell_summary <- cube_tbl |>
  filter(!is.na(pm2_5)) |>
  summarise(pm2_5_mean = mean(pm2_5), n_years = n(), .by = c(x, y)) |>
  tibble_to_cube(cube)
```

## Adding a collection

Unknown collections fall back to the generic raster reader, which usually works.
To pin down the file to pick, the attribute name and the resampling method, add a
spec:

```{r}
specs <- c(
  geo_admin_specs(),
  list("ch.example.collection" = collection_spec(
    read_collection_rasters,
    label      = "example",
    resampling = "average",
    name       = "example",
    formats    = "tif",
    pattern    = "_(19|20)\\d{2}_2056\\.tif$"
  ))
)

read_geo_admin("ch.example.collection", specs = specs)
```
