---
title: "Polar plots: concentration by wind direction and speed"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polar plots: concentration by wind direction and speed}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_openair <- requireNamespace("openair", quietly = TRUE)
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 5.5, fig.height = 4.5,
                      eval = has_openair)
```

```{r}
#| message: false
#| warning: false
library(airquality.methods)
library(ggplot2)
library(openair)
```

A polar plot shows a concentration as a function of wind direction (angle) and wind
speed (radius). `openair::polarPlot()` computes a smoothed surface and draws it with
lattice; the functions here draw such surfaces — or unsmoothed cells binned straight
from the data — as ggplot2 objects, so they can be themed, faceted, combined and
placed on a map (`vignette("polar-maps")`).

Two design decisions run through everything below:

* The rose is drawn on a **Cartesian** panel with `coord_fixed()`, not with
  `coord_polar()`. The cells already live in the wind-vector plane (u, v), and
  `coord_polar()` would warp them.
* Therefore the rings and spokes are layers, styled through a theme element of their
  own, `polar.grid` — ggplot2's `panel.grid` would draw *straight* lines through the
  rose.

The examples use openair's `mydata` (Marylebone Road, London), which is why this
vignette needs the openair package.

```{r}
d <- selectByDate(mydata, year = c(2003, 2004))
```

## The two entry points

One call that computes and draws: `...` goes to `openair::polarPlot()`, `args` to
`polar_raster()`.

```{r}
d |> polar_plot(pollutant = "nox")
```

Two steps — compute first, then draw. Worth it once the display is varied repeatedly,
because the smoothing is the expensive part.

```{r}
pp <- polarPlot(d, pollutant = "nox", plot = FALSE)
polar_raster(pp)
```

## Why the display arguments live in `args`

`polarPlot()` already claims the names `limits`, `type` and `x`, each with a different
meaning than here. A shared `...` would silently route them to the wrong place — hence
the split:

```{r}
#| fig.width: 7
d |> polar_plot(
  pollutant = "nox",                       # -> polarPlot()
  type      = "season",                    # -> polarPlot() (facets)
  args      = list(limits = c(50, 250),    # -> colour scale
                   name   = "NOx")
)
```

A typo in `args` is reported instead of silently swallowed:

```{r}
#| error: true
d |> polar_plot(pollutant = "nox", args = list(limmits = c(50, 250)))
```

## The colour scale

The default is `scale_fill_viridis_c(option = "A")` (magma, a dark-to-light ramp that
stays apart from the greens of a topographic basemap); `...` goes there.

```{r}
polar_raster(pp, name = "NOx")
polar_raster(pp, limits = c(50, 250), name = "NOx")   # values outside -> grey
polar_raster(pp, option = "A", direction = -1, name = "NOx")
polar_raster(pp, trans = "log10", name = "NOx")
```

Any other ggplot2 fill scale can be supplied:

```{r}
polar_raster(pp, scale = scale_fill_distiller(palette = "RdYlBu", name = "NOx"))
polar_raster(pp, scale = scale_fill_steps(low = "#FFFFCC", high = "#800026",
                                          n.breaks = 6, name = "NOx"))
```

## The capped scale

Capping is not built in, it composes from outside (`vignette("capped-scales")`). Two
equivalent ways — the first stays silent, the second reports the usual "Scale for fill
is already present".

```{r}
polar_raster(pp, scale = scale_fill_capped(limits = c(50, 250), name = "NOx"))
polar_raster(pp) + scale_fill_capped(limits = c(50, 250), name = "NOx")
```

Cap only the top, take the limits from quantiles of the grid values, or bin:

```{r}
polar_raster(pp, scale = scale_fill_capped(limits = c(NA, 200), name = "NOx"))
polar_raster(pp, scale = scale_fill_capped(probs = c(0.1, 0.9), name = "NOx"))
polar_raster(pp, scale = scale_fill_capped_b(limits = c(50, 250), n.breaks = 5,
                                             palette = "YlOrRd", direction = -1))
```

Inside the one-call wrapper the scale is an element of `args`:

```{r}
#| fig.width: 7
d |> polar_plot(pollutant = "nox", type = "season",
                args = list(scale = scale_fill_capped(limits = c(50, 250),
                                                      name = "NOx")))
```

## Facets

`polarPlot()`'s grouping column is named differently depending on the call
(`default`, `season`, `variable`, ...) and is detected automatically. All facets share
one colour scale.

```{r}
#| fig.width: 7
pp_season <- polarPlot(d, pollutant = "nox", type = "season", plot = FALSE)
polar_raster(pp_season, limits = c(50, 250), name = "NOx")
```

Two grouping columns at once. With pollutants at very different levels the weaker
panels turn flat — one call per pollutant is then the better choice.

```{r}
#| fig.width: 7
pp_two <- polarPlot(d, pollutant = c("nox", "so2"), type = "weekend", plot = FALSE)
polar_raster(pp_two)
```

Faceting off (all groups on top of each other), or named explicitly:

```{r}
#| eval: false
polar_raster(pp_season, facet = NA)
polar_raster(pp_season, facet = ~season)
```

## The polar grid

Everything that is not the surface is set in `polar_grid()`.

```{r}
polar_raster(pp, grid = polar_grid(rings = c(3, 6, 9, 12), spokes = seq(0, 330, 30)))
polar_raster(pp, grid = polar_grid(compass = "8"))                      # eight points
polar_raster(pp, grid = polar_grid(compass = c(Luv = 225, Lee = 45)))   # own marks
```

The grid below the surface (visible only outside the data lobe), a more restrained
grid, and more margin for the compass letters:

```{r}
polar_raster(pp, grid = polar_grid(below = TRUE))
polar_raster(pp, grid = polar_grid(outer = FALSE, spokes = NULL, colour = "grey70"))
polar_raster(pp, grid = polar_grid(expand = 0.3, compass_size = 4))
```

The grid is themed, not just parameterised. `polar_grid()` reads the theme element
`polar.grid` whenever `colour`, `linewidth` or `linetype` are left at `NULL`:

```{r}
polar_raster(pp, theme = theme_polar(grid_colour = "grey70", grid_linetype = 1))

polar_raster(pp, theme = theme_polar() +
                   theme(polar.grid = element_line(colour = "white", linetype = 1,
                                                   linewidth = 0.2)))
```

An explicit argument beats the theme, field by field — this is grey20, dashed, 0.25:

```{r}
polar_raster(pp, grid = polar_grid(colour = "grey20"),
             theme = theme_polar(grid_colour = "white", grid_linewidth = 1))
```

One caveat: the style is read when the plot is *built*, so the theme has to arrive
through `theme =`. A `+ theme(polar.grid = ...)` added afterwards comes too late — the
layers already exist.

## The radial axis

It is the plot's ordinary y-axis: the title via `axis_title`, everything else via
`theme()`. `guide_axis(cap = "both")` trims the line to the range up to the largest
ring. The unit sits on every break by default (`axis_unit = "m/s"`), so there is no
axis title unless asked for.

```{r}
polar_raster(pp, axis_title = "wind speed (m/s)", axis_unit = NULL)

polar_raster(pp) +
  theme(axis.text.y  = element_text(colour = "firebrick", face = "bold", size = 11),
        axis.ticks.y = element_line(colour = "firebrick", linewidth = 0.6),
        axis.line.y  = element_line(colour = "firebrick", linewidth = 0.6))
```

Another unit, none at all, or no axis:

```{r}
polar_raster(pp, axis_unit = "km/h")
polar_raster(pp, axis_labels = FALSE)
```

Or moved into the panel, from the centre outwards. ggplot2 draws axes in the gtable
*around* the panel and nowhere else, so this is no longer an axis in ggplot's sense but
one layer: a label per ring, sitting on its ring. No axis line and no ticks — the rings
already are the ticks. What survives is the rest: the same rings as the grid, the same
break labels as the key rose of `polar_map()`, and `axis.text.y` for the styling. What
is gone is `guide_axis()` and `axis_title`.

```{r}
polar_raster(pp, axis_labels = "centre", theme = theme_polar(axis_colour = "grey10"))
polar_raster(pp, axis_labels = "centre", axis_bearing = 135)
```

The centred axis is the better choice when the figure is small or the left margin is
needed elsewhere — and it is what the key rose of `polar_map()` does, for the same
reason: there, the y-axis is the northing.

## Theme and labelling

```{r}
polar_raster(pp, theme = theme_polar(base_size = 14)) +
  labs(title = "NOx by wind direction and speed")
polar_raster(pp) + theme(legend.position = "bottom")
```

## Recolouring without recomputing

`polar_plot()` carries the grid data as an attribute and `polar_data()` retrieves it,
which saves the smoothing step.

```{r}
#| fig.width: 7
p <- d |> polar_plot(pollutant = "nox", type = "season",
                     args = list(limits = c(50, 250)))

polar_raster(polar_data(p), option = "C")
polar_raster(polar_data(p), scale = scale_fill_capped(probs = 0.02),
             grid = polar_grid(compass = "8"))
```

## Other polarPlot() statistics

Anything `polarPlot()` delivers as a u/v/z grid can be drawn, and its smoothing
arguments (`k`, ...) go through `...`.

```{r}
d |> polar_plot(pollutant = "so2", statistic = "cpf", percentile = 90,
                args = list(limits = c(0.1, 0.5), name = "CPF", option = "B"))
d |> polar_plot(pollutant = "nox", k = 50)
```

## Without smoothing: `polar_bin()`

Instead of smoothing a surface, the measurements are sorted into square cells of the
wind-vector plane. Nothing is interpolated — empty cells stay empty. The result goes
straight into `polar_raster()`.

```{r}
d |> polar_bin("nox", res = 2) |> polar_raster(name = "NOx")
```

On choosing `res`: wind directions here come in 10-degree steps. Two adjacent
directions are about r · 10 · π/180 apart — at r = 10 that is 1.75. A finer grid leaves
gaps there and the image frays into rays. That is not a bug, it is the resolution of the
raw data:

```{r}
d |> polar_bin("nox", res = 0.5) |> polar_raster(name = "NOx")   # fine inside, rays outside
```

Thinly populated cells are dropped with `min_n` — otherwise single measurements show up
as strong blotches.

```{r}
d |> polar_bin("nox", res = 1, min_n = 10) |> polar_raster(name = "NOx")
```

Other statistics; `"n"` shows the data density itself, and any function works:

```{r}
d |> polar_bin("nox", res = 1, statistic = "median", min_n = 5) |>
  polar_raster(name = "Median")
d |> polar_bin("nox", res = 1, statistic = "n") |>
  polar_raster(name = "Count", trans = "log10")
d |> polar_bin("nox", res = 1, min_n = 20,
               statistic = function(x) quantile(x, 0.95)) |>
  polar_raster(name = "P95")
```

`polar_statfun()` turns a `statistic` into the function the binning uses, so an
analysis that condenses values per direction on its own can mean exactly the same by
`"median"`.

Grouping as with `polarPlot()` — derived types go through `openair::cutData()` — and a
limited radius with a capped scale:

```{r}
#| fig.width: 7
d |> polar_bin("nox", res = 1, min_n = 5, type = "season") |>
  polar_raster(name = "NOx")
```

```{r}
d |> polar_bin("nox", res = 1, min_n = 5, ws_max = 8) |>
  polar_raster(scale = scale_fill_capped(limits = c(50, 250), name = "NOx"))
```

Binning rounds half away from zero (`round_off()`), never with `round()`. Wind data come
in fixed steps, so a value on a cell boundary is not a rare accident but half the
dataset, and round-half-to-even would give one cell of every pair three times the
observations of the other.

`polar_raster()` draws sharp cells by default: invented smoothness would be misleading
for aggregated cells. Interpolation can be forced:

```{r}
d |> polar_bin("nox", res = 1) |> polar_raster(interpolate = TRUE)
```

## Sectors instead of squares: `polar_sector()`

Here the cells are ring segments of wind direction and speed class. They follow the
structure of the data and therefore leave no gaps — unlike the square grid.

```{r}
d |> polar_sector("nox", wd_res = 10, ws_res = 1) |> polar_raster(name = "NOx")
d |> polar_sector("nox", wd_res = 30, ws_res = 2) |> polar_raster(name = "NOx")
```

`wd_res` must divide 360 (5, 10, 15, 22.5, 30, 45, ...). The statistics, facets and
scales work as for `polar_bin()`:

```{r}
d |> polar_sector("nox", wd_res = 22.5, ws_res = 1.5, min_n = 10) |>
  polar_raster(name = "NOx")
d |> polar_sector("nox", statistic = "n", min_n = 20) |>
  polar_raster(name = "Count", trans = "log10")
d |> polar_sector("nox", min_n = 10) |>
  polar_raster(scale = scale_fill_capped(limits = c(50, 250), name = "NOx"))
```

## Any u/v grid

The input does not have to be an openair object — a data frame with u, v and a value is
enough, and the column names are free.

```{r}
grid_df <- as.data.frame(pp$data)
names(grid_df)[names(grid_df) == "z"] <- "conc"
polar_raster(grid_df, z = "conc", limits = c(50, 250))
```

## Saving

The surface is a raster, not a cloud of points — in vector output roughly fourteen
times smaller than openair's own plot.

```{r}
#| eval: false
ggsave("polar_nox.pdf", polar_raster(pp, limits = c(50, 250)),
       width = 6, height = 5, device = grDevices::cairo_pdf)
ggsave("polar_nox.png", polar_raster(pp, limits = c(50, 250)),
       width = 6, height = 5, dpi = 300)
```

## One pollutant, several statistics side by side

"Which statistic" is not a data column, so it cannot be a `type`. Instead, call once per
statistic, label each result, and bind the rows; `polar_raster()` then facets on the
label like on any other grouping column.

```{r}
#| fig.width: 7
#| message: false
metrics <- dplyr::bind_rows(
  polarPlot(d, pollutant = "nox", statistic = "mean",
            plot = FALSE)$data |> dplyr::mutate(metric = "mean"),
  polarPlot(d, pollutant = "nox", statistic = "median",
            plot = FALSE)$data |> dplyr::mutate(metric = "median"),
  polarPlot(d, pollutant = "nox", statistic = "percentile", percentile = 90,
            plot = FALSE, auto.text = FALSE)$data |> dplyr::mutate(metric = "p90")
)
polar_raster(metrics, name = "NOx")
```

The same pattern works for unsmoothed cells:

```{r}
#| fig.width: 7
metrics_bin <- dplyr::bind_rows(
  polar_bin(d, "nox", res = 1, min_n = 5, statistic = "mean") |>
    dplyr::mutate(metric = "mean"),
  polar_bin(d, "nox", res = 1, min_n = 5, statistic = "median") |>
    dplyr::mutate(metric = "median")
)
polar_raster(metrics_bin, name = "NOx")
```

## A key rose for a facetted figure: `polar_key()`

A panel of many roses has a problem a single rose does not: compass letters and radial
labels are the same on every panel, and they are drawn over the very cells one wants to
read. On a map, `polar_map()` solves this with one key rose off to the side;
`polar_key()` is that key as a plot of its own.

It cannot describe something else: it shares the rings and grid layers with
`polar_raster()` and its ring labels with the key rose of `polar_map()`. Pass the data
the panels are drawn from — the outer radius comes from it (a number works too).

```{r}
#| fig.width: 8
#| fig.height: 5
panels <- polar_raster(pp_season, grid = polar_grid(compass = FALSE),
                       axis_labels = "none", name = "NOx")
key <- polar_key(pp_season)

if (requireNamespace("patchwork", quietly = TRUE)) {
  patchwork::wrap_plots(panels, key, widths = c(2, 1))
}
```

The compass can be lettered in any language, and `title` gives the key a heading:

```{r}
#| fig.width: 3
#| fig.height: 3
polar_key(pp_season, compass = c(N = 0, O = 90, S = 180, W = 270), title = "Wind")
polar_key(6, compass = "8", axis_unit = NULL, n_labels = Inf)
```
