---
title: "Polar maps: wind roses at their true place, on a swisstopo basemap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Polar maps: wind roses at their true place, on a swisstopo basemap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
has_openair <- requireNamespace("openair", quietly = TRUE)
# The basemap is fetched from the swisstopo web map service; without network
# (or without png) those chunks are shown but not run.
has_basemap <- has_openair &&
  requireNamespace("png", quietly = TRUE) &&
  requireNamespace("curl", quietly = TRUE) && curl::has_internet()
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6.5, fig.height = 6,
                      eval = has_openair)
```

```{r}
#| message: false
#| warning: false
library(airquality.methods)
library(ggplot2)
library(openair)
library(dplyr)
```

`polar_map()` places polar plots (`vignette("polar-plots")`) at the true LV95
coordinates of their sites, optionally over a swisstopo basemap. It is meant for a
monitoring network: one rose per site, all on one ruler, so a ring is the same wind
speed everywhere and a single key rose describes them all.

## Test data

The site coordinates are those of eight sites of an ultrafine particle network in
Kloten (published OGD metadata), because a basemap needs real LV95 metres to sit under.
The measurements are openair's `mydata` (Marylebone Road, London), one year per site —
eight genuinely different roses that say nothing about Kloten, but run from a package
dataset.

**Careful with the column names.** Here `x_lv95` is the **easting** and `y_lv95` the
**northing**; the letters invert the Swiss E/N convention. `polar_map()` does not guess
column names for exactly this reason — `east` and `north` name them, and a transposed
pair is an error, not a map of the wrong place.

```{r}
sites <- data.frame(
  site   = c("B1", "A4", "C0", "B2", "B3", "A1", "A3", "A2"),
  x_lv95 = c(2687216.0, 2686274.5, 2688846.2, 2686979.8,
             2686836.8, 2686115.0, 2686371.0, 2686332.5),
  y_lv95 = c(1257939.9, 1256170.2, 1257367.6, 1256701.4,
             1255717.6, 1257708.3, 1256638.8, 1257197.6)
)

d <- bind_rows(Map(
  function(s, yr) transform(
    selectByDate(mydata, year = yr)[, c("date", "ws", "wd", "nox")], site = s),
  sites$site, 1998:2005
))
```

## The minimal call

`x` is one long data frame with a `site` column plus the usual u/v/z — the same
"bind the rows, label them" idiom as for several statistics in `vignette("polar-plots")`,
here the label is a place. A named list works directly too.

```{r}
per <- lapply(split(d, d$site), polar_bin, "nox", res = 1, min_n = 5)
polar_map(per, sites)

m <- bind_rows(per, .id = "site")
p <- polar_map(m, sites)
```

`polar_map()` reports the radius it chose and why: the largest at which no two roses
touch, here set by the two closest sites (A3 and A4, 478 m apart), with the compass
margin of the grid included.

## `radius`: how big is a rose

`radius` is the outer ring in metres on the ground. The drawn footprint is
`radius * (1 + grid$expand)`, because the compass letters sit out in that margin — that
factor is part of the arithmetic.

```{r}
polar_map(m, sites, radius = 150)
```

Asking for more than fits is allowed, and says so. On a regional overview that is often
the right trade; `suppressWarnings()` if you mean it.

```{r}
polar_map(m, sites, radius = 400)
```

## One ruler for every rose

`k = radius / r_max` is a single number for the whole map. A given ring is the same wind
speed at every site, which is what makes the map comparable — and what lets one key rose
describe it. So if one site has a long tail, do **not** rescale per site; clip all of
them:

```{r}
polar_map(m, sites, r_max = 6, quiet = TRUE)     # everything above 6 m/s is dropped
```

Clipping drops whole cells, never rows: for sector data one row is a polygon vertex,
and clipping row by row would leave slivers that are in the data nowhere.

The scale is recorded on the plot, for annotations or checks:

```{r}
attr(p, "polar_map_scale")         # k (m per m/s), radius, r_out
```

## The basemap

`bbox_lv95()` derives the extent from the sites (padded by the rose radius, rounded
outward to whole metres); `basemap_swisstopo()` fetches one georeferenced PNG from the
swisstopo web map service and caches it on disk.

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
bb <- bbox_lv95(sites, radius = 200)
bm <- basemap_swisstopo(bb, px = 1000)
bm                                   # layer, extent, size, m/px, file

polar_map(m, sites, basemap = bm, quiet = TRUE)
```

Resolution: `px` is pixels along the longer edge, `res` is metres per pixel. The other
edge follows from the box, so the image is never stretched. There is deliberately no
`dpi` argument — dpi means nothing without a physical figure size, which the server
does not know.

```{r}
#| eval: false
basemap_swisstopo(bb, px = 3000)
basemap_swisstopo(bb, res = 1)       # 1 m/px
```

`bm$asp` is the aspect of the map; size the figure by it and nothing is letterboxed:

```{r}
#| eval: false
ggsave("map.png", polar_map(m, sites, basemap = bm, quiet = TRUE),
       width = 9, height = 9 / bm$asp + 0.9, dpi = 200, bg = "white")
```

Any named vector with `xmin`, `ymin`, `xmax`, `ymax` in LV95 works as a box —
`sf::st_bbox()` of an LV95 object, or `bbox_zh_lv95` for the whole canton.

## Which basemap layer

`"colour"` (default) is the Landeskarte: it reads as a map, which is what a reader needs
to place a rose in a town. `"grey"` is the quieter choice when the roses have to carry
the figure alone — the colour map puts saturated forest green and motorway orange right
where a sequential fill needs to be read. The magma default of the roses keeps its
distance from those greens better than viridis would. `"image"` is aerial imagery.

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
polar_map(m, sites, basemap = basemap_swisstopo(bb, layer = "grey", px = 1000, quiet = TRUE),
          quiet = TRUE)
polar_map(m, sites, basemap = basemap_swisstopo(bb, layer = "image", px = 1000, quiet = TRUE),
          quiet = TRUE)
```

Any other WMS layer id is passed through untouched:

```{r}
#| eval: false
basemap_swisstopo(bb, layer = "ch.swisstopo.pixelkarte-farbe-pk25.noscale")
```

## Pinning the basemap for a Quarto report

`file` writes the PNG plus a `.pgw` world file, a `.prj` and a `.layer` beside it. Once
that file exists it is read directly and no request is made — the first render populates
it, and every later render (CI, a colleague's laptop, offline) reuses it byte for byte.
The world file also makes it open correctly in QGIS.

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
pin <- file.path(tempdir(), "basemap_example.png")
bm  <- basemap_swisstopo(bb, px = 1000, file = pin, quiet = TRUE)
list.files(tempdir(), pattern = "^basemap_example")
```

A pinned file wins over `layer`, `px` and `res`, and warns when they disagree with it.
Caching is not just convenience: swisstopo's terms ask that the service not be queried
at high intensity, and a report that re-renders often would. Refresh only when the
extent or layer really changed:

```{r}
#| eval: false
basemap_swisstopo(bb, file = pin, refresh = TRUE)
```

## The key rose and the scale bar

The map has two rulers, and they measure different things: the rose radius is wind
speed, map distance is metres. Both are labelled. The key rose is drawn with the same
`k` as the data roses, so its rings are physically identical to theirs — not kept in
sync, the same number used twice.

```{r}
polar_map(m, sites, decor = polar_map_decor(key = "topright"), quiet = TRUE)
```

It carries no heading by default, for the same reason the radial axis has no title: the
unit is on every ring label already. Add one if the figure is read out of context.

```{r}
polar_map(m, sites, decor = polar_map_decor(key_title = "Windgeschwindigkeit"),
          quiet = TRUE)
```

If the key lands on a site, `polar_map()` says so and names the free corners. It can be
placed by hand (map coordinates), or dropped; the scale bar length is a round fraction of
the map width unless set:

```{r}
#| eval: false
polar_map(m, sites, decor = polar_map_decor(key = c(2688940, 1255850)))
polar_map(m, sites, decor = polar_map_decor(key = FALSE))
polar_map(m, sites, decor = polar_map_decor(scalebar = FALSE))
```

```{r}
polar_map(m, sites, decor = polar_map_decor(scalebar_length = 1000), quiet = TRUE)
```

Site labels and markers, and LV95 coordinates on the axes (off by default — the scale
bar carries the distance, and the numbers are noise for most readers):

```{r}
polar_map(m, sites, decor = polar_map_decor(label_size = 2.2, label_gap = 0.3),
          theme = theme_polar_map(axes = TRUE), quiet = TRUE)
```

## The polar grid on a map

The same `polar_grid()` as in `polar_raster()`, and the same theme element behind it:
`theme_polar_map()` sets `polar.grid` darker and slightly heavier than `theme_polar()`,
because a plain-background grey disappears over a basemap.

The compass is off on the roses by default here: eight sites times four letters is
clutter, and every rose is oriented alike, so the key rose says it once. It can be turned
back on per rose, and the key's compass is a setting of its own:

```{r}
polar_map(m, sites, grid = polar_grid(compass = TRUE), quiet = TRUE)
polar_map(m, sites, decor = polar_map_decor(key_compass = "8"), quiet = TRUE)
```

Restyle the grid the theme way (through `theme =`, not `+ theme()`), and fix the rings so
the key reads in round numbers:

```{r}
polar_map(m, sites, grid = polar_grid(rings = c(2, 4, 6, 8)),
          theme = theme_polar_map(grid_colour = "grey10", grid_linetype = 1),
          quiet = TRUE)
```

## Colour scales

`...` goes to the default fill scale, exactly as in `polar_raster()`; the capped scale
composes as usual (`vignette("capped-scales")`).

```{r}
polar_map(m, sites, limits = c(50, 400), name = "NOx (ppb)", quiet = TRUE)
polar_map(m, sites, scale = scale_fill_capped(limits = c(50, 400), name = "NOx (ppb)"),
          quiet = TRUE)
```

## Sectors instead of squares

On a map, sectors are usually the better choice: they follow the structure of the data,
so they leave no gaps for the basemap to shine through the middle of a rose, and they stay
one polygon layer whatever the number of sites.

```{r}
sec <- bind_rows(lapply(split(d, d$site), polar_sector, "nox",
                        wd_res = 15, ws_res = 1, min_n = 5), .id = "site")
polar_map(sec, sites, quiet = TRUE)
```

A note on cell counts: a default `polarPlot()` grid has about 10 000 cells, and eight of
them at 200 m radius on a 3.5 km map render at about 2 px per cell. The smoothing detail
is invisible and you pay 80 000 cells for it. `polar_bin(res = 1)` or
`polar_sector(wd_res = 15)` is the honest resolution.

## A regional overview is a separate figure

Add a ninth site 7 km to the east (and a ninth rose, from all of `mydata`):

```{r}
sites_all <- bind_rows(sites, data.frame(site = "D0", x_lv95 = 2693324, y_lv95 = 1258946))
m_all <- bind_rows(m, transform(polar_bin(mydata, "nox", res = 1, min_n = 5), site = "D0"))
```

The problem is scale, not code. The cluster still pins the default radius at about
200 m, but the map is now 7 km wide, so a rose is under 3 % of it — invisible. Two honest
options: a representative subset at a radius that reads, or everything at a radius that
overlaps in the cluster — knowingly.

```{r}
polar_map(filter(m_all, site %in% c("A3", "D0")),
          filter(sites_all, site %in% c("A3", "D0")), radius = 800, quiet = TRUE)

suppressWarnings(
  polar_map(m_all, sites_all, radius = 500, decor = polar_map_decor(labels = FALSE),
            quiet = TRUE)
)
```

What does not work is one figure that is both. Make the detail map and the overview, and
put them side by side.

## The detail map, final form

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
p <- polar_map(
  sec, sites,
  basemap = bm,
  scale   = scale_fill_capped(limits = c(50, 400), name = "NOx (ppb)"),
  decor   = polar_map_decor(key = "topright"),
  quiet   = TRUE
)
p
```

In a Quarto report, the title and caption belong to the figure object rather than into
the panel — `fig_meta()` attaches them without drawing them:

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
p <- fig_meta(p, title = "NOx by wind direction and speed",
              caption = "openair::mydata, one year per site.")
fig_caption(p)
```

## Attribution

swisstopo geodata are free to use **with attribution**. `polar_map()` adds
"Kartengrundlage: © swisstopo" as the caption whenever a basemap is drawn; do not remove
it from published figures. It comes from the basemap object, not from the decor, so a
map without basemap carries no caption — there is nothing to attribute.

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
bm$attribution
polar_map(m, sites, basemap = bm, quiet = TRUE)$labels$caption
is.null(polar_map(m, sites, quiet = TRUE)$labels$caption)
```

It can be reworded (an English report, say), but keep the source:

```{r}
#| eval: false
polar_map(m, sites, basemap = bm,
          decor = polar_map_decor(attribution = "Map data: © swisstopo"))
```

## A basemap and a scale bar under any LV95 ggplot

`annotation_basemap()` and `annotation_scalebar()` work on their own — `polar_map()` is
not required. Limit the panel to the same box the basemap covers, and the scale bar
cannot drift out of it; `polar_map()` draws its scale bar with the very same function.

```{r}
#| eval: !expr has_basemap
#| dev: jpeg
ggplot(sites, aes(x_lv95, y_lv95)) +
  annotation_basemap(bm) +
  geom_point(size = 3, colour = "firebrick") +
  geom_text(aes(label = site), nudge_y = 90, size = 3) +
  coord_fixed(xlim = bb[c("xmin", "xmax")], ylim = bb[c("ymin", "ymax")],
              expand = FALSE) +
  annotation_scalebar(bb) +
  theme_polar_map() +
  labs(caption = bm$attribution)
```
