---
title: "Colour scales for air quality figures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Colour scales for air quality figures}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 4)
```

```{r}
#| message: false
library(airquality.methods)
library(ggplot2)
```

## The problem with limits

A concentration map needs fixed limits: the same pollutant must look the same
across years and municipalities, or the colours say nothing. But fixed limits
mean some values fall outside them, and ggplot2's default answer is to turn those
cells into `NA` — holes in the map, exactly where the pollution is worst.

`scale_capped()` squishes out-of-range values onto the edge colour instead, marks
that in the legend with "≤"/"≥", and draws the true data extremes as whiskers
beyond the colour bar. Nothing is hidden and nothing is lost.

```{r}
d <- data.frame(expand.grid(x = 1:20, y = 1:20))
d$value <- as.vector(outer(1:20, 1:20, \(a, b) a * b / 4))

ggplot(d, aes(x, y, fill = value)) +
  geom_raster() +
  scale_fill_capped(limits = c(10, 60), name = "value") +
  theme_custom("map")
```

The legend shows `≥ 60` and a whisker at the real maximum, so a reader can see
both the classification and what it cost.

## Pollutant scales

`immissionscale()` supplies the limits, breaks and legend title agreed for each
parameter, so figures are comparable by construction.

```{r}
immissionscale_parameters()
```

```{r}
#| fig.height: 3
pm <- data.frame(expand.grid(x = 1:30, y = 1:10))
pm$value <- seq(0, 25, length.out = nrow(pm))

ggplot(pm, aes(x, y, fill = value)) +
  geom_raster() +
  immissionscale("PM2.5") +
  coord_equal() +
  theme_custom("map")
```

Concentrations use the 18-step official ramp; nitrogen (`NH3`, `Ndep`) uses a
reversed magma ramp, where dark means high.

## Choosing limits from the data

When sensible limits are not known in advance, let quantiles set them — the
capping then trims the tail instead of the colour ramp being dictated by a single
outlier:

```{r}
ggplot(d, aes(x, y, fill = value)) +
  geom_raster() +
  scale_fill_capped(probs = 0.02, name = "value") +
  theme_custom("map")
```

`probs = 0.02` is shorthand for `c(0.02, 0.98)`. `probs = c(NA, 0.98)` caps only
the upper end.

## Classed maps

For figures that report classes rather than a continuum:

```{r}
ggplot(d, aes(x, y, fill = value)) +
  geom_raster() +
  scale_fill_capped_b(limits = c(10, 60), breaks = seq(10, 60, 10), name = "value") +
  theme_custom("map")
```

`scale_colour_capped()` and the `_b` / `_d` variants cover the colour aesthetic
and the binned and discrete scale types.

## Turning the capping off

Passing a different `oob` makes it an ordinary scale again — cap symbols and
whiskers then disappear on their own, because there is no edge class left to
mark:

```{r}
#| fig.height: 3
ggplot(d, aes(x, y, fill = value)) +
  geom_raster() +
  scale_fill_capped(limits = c(10, 60), oob = scales::oob_censor, name = "value") +
  theme_custom("map")
```

## Themes

`theme_custom()` strips what a figure type does not need:

```{r}
#| fig.height: 3
ggplot(economics, aes(date, unemploy)) +
  geom_line() +
  theme_custom("timeseries")
```

`"timeseries"`, `"scatter"`, `"flipped"`, `"map"` and `"default"`.
`theme_legend_inside()` moves the legend into the panel, which maps usually have
room for:

```{r}
#| fig.height: 3
ggplot(pm, aes(x, y, fill = value)) +
  geom_raster() +
  immissionscale("PM2.5") +
  coord_equal() +
  theme_custom("map") +
  theme_legend_inside()
```
