---
title: "STATPOP: the hectare grid and its collector pixels"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{STATPOP: the hectare grid and its collector pixels}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| include: false
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
```

```{r}
#| message: false
library(airquality.methods)
library(dplyr)
```

BFS publishes the resident population per hectare as part of the STAC collection
`ch.bfs.statistik-bevoelkerung_haushalte`. Two details decide whether the numbers
that come out are right.

## Detail 1: the coordinates mark the corner, not the centre

A STATPOP row carries `E_KOORD` / `N_KOORD` for the **lower left corner** of its
cell. Treating those as cell centres shifts the whole population half a cell
south-west — 50 m, which is enough to move people across a municipal boundary.

`table_to_stars()` takes the reference point as an argument and, more usefully,
refuses to guess: it checks that every coordinate really lies on the grid.

```{r}
cells <- tibble(
  E_KOORD = c(2600000, 2600100, 2600200),
  N_KOORD = c(1200000, 1200000, 1200000),
  BBTOT   = c(10, 20, 30)
)

raster <- table_to_stars(cells, variables = "BBTOT", cellsize = 100)
sf::st_bbox(raster)
```

The extent starts exactly at the first corner and ends one cell beyond the last.
Feed it something off the grid and it says so instead of snapping:

```{r}
#| error: true
table_to_stars(
  tibble(E_KOORD = c(2600000, 2600042), N_KOORD = 1200000, BBTOT = c(1, 2)),
  variables = "BBTOT", cellsize = 100
)
```

## Detail 2: the collector pixels

Every municipality has one cell that absorbs the inhabitants who cannot be
located. Left in place, those people sit at a single arbitrary coordinate and get
assigned whatever pollutant concentration happens to be there — a small number of
cells with implausibly high population and an exposure that means nothing.

BFS ships them separately (`_noloc_`), and `subtract_noloc()` removes them from
the cell they were added to, logging every one:

```{r}
population <- tibble(
  E_KOORD = rep(c(2600000, 2600100), each = 2),
  N_KOORD = rep(c(1200000, 1200100), times = 2),
  BBTOT   = c(10, 20, 30, 40)
) |>
  table_to_stars(variables = "BBTOT", cellsize = 100)

noloc <- tibble(E_KOORD = 2600100, N_KOORD = 1200000, BBTOT = 5)

corrected <- subtract_noloc(population, noloc)
corrected$noloc
```

```{r}
sum(population$BBTOT, na.rm = TRUE)         # before
sum(corrected$stars$BBTOT, na.rm = TRUE)    # after
```

### The four outcomes

The log names what happened, per pixel:

| status | meaning |
|---|---|
| `subtracted` | removed in full |
| `exceeds_cell` | more to remove than the cell held; the cell went to 0 |
| `cell_missing` | the cell is empty (`NA`); nothing removed |
| `outside_extent` | the pixel lies outside the bbox; nothing removed |

The last one is routine, not an error: cropping to the Canton of Zurich leaves
the collector pixels of every other canton outside. They are still counted, so
the books balance:

```
sum(noloc file) = noloc_subtracted + noloc_unmatched
```

The first three warn, because they mean the correction did not fully apply.

## Reading a year

```{r}
#| eval: false
assets  <- get_geo_admin_assets("ch.bfs.statistik-bevoelkerung_haushalte")
statpop <- read_statpop_ha(assets, years = 2015:2025)

statpop |> select(year, format, noloc_subtracted, noloc_unmatched)
statpop$noloc[[1]] # the per-pixel log
```

The file hierarchy is resolved automatically: `_ha_` selects the hectare grid
over the municipality table, parquet is preferred over csv, and EPSG 2056 over
the rest.

### Checking the result

Localised plus non-localised inhabitants should reproduce the published
population of the extent:

```{r}
#| eval: false
statpop |>
  mutate(
    in_raster = purrr::map_dbl(stars, \(s) sum(s$BBTOT, na.rm = TRUE)),
    total     = in_raster + noloc_subtracted
  ) |>
  select(year, in_raster, noloc_subtracted, total)
```

## Reading more than the headcount

The hectare grid holds far more than `BBTOT` — age classes (`BBM*`, `BBW*`),
household types (`HP*`), nationality and civil status groups. Name them, and mind
the memory: Switzerland at 100 m is about 8 million cells per variable.

```{r}
#| eval: false
read_statpop_ha(
  assets,
  years     = 2024,
  variables = c("BBTOT", "BB11", "BB12"),
  bbox      = bbox_zh_lv95
)
```

Only `noloc_variable` (`BBTOT` by default) is corrected for collector pixels;
the other variables come through as published, and you are told so.

## Switching the correction off

Sometimes the uncorrected raster is what is wanted, for a comparison against
published figures:

```{r}
#| eval: false
read_statpop_ha(assets, years = 2024, correct_noloc = FALSE)
```

The `noloc_*` columns are then `NA` rather than 0, so a corrected and an
uncorrected result can never be confused with one another.
